Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-05-30 iso-int/rc-iso-int.tex
fixes
amr done

\newcommand{\todo}[1]{\textbf{TODO: #1}}

@@

\begin{document}
\title{Isomorphic Interpreters from Logically Reversible Abstract Machines} 
\titlerunning{On the construction of Isomorphic Interpreters} 
\author{Roshan P. James \and Amr Sabry}
\institute{School of Informatics and Computing, Indiana University\\
\mails}
\maketitle

@@

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Axelsen:2011:RPC:1987171.1987176,Yokoyama:2008:PRP,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of transforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

\paragraph*{Graphical Language.}
Following the tradition established for monoidal
categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
graphical language that conveys the semantics of the language.  The general
idea of the graphical notation is that combinators are modeled by ``wiring
diagrams'' or ``circuits'' and that values are modeled as ``particles'' or
``waves'' that may appear on the wires. Evaluation therefore is modeled by
the flow of waves and particles along the wires.

\begin{itemize}
\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
  is simply represented as a wire labeled by its type {{b}}, as shown on the
  left. In more complex diagrams, if the type of a wire is obvious from the
  context, it may be omitted. When tracing a computation, one might imagine a
  value {{v}} of type {{b}} on the wire, as shown on the right.

@@

\item The product type {{b1*b2}} may be represented using either one wire
  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
  case of products represented by a pair of wires, when tracing execution
  using particles, one should think of one particle on each wire or
  alternatively as in folklore in the literature on monoidal categories as a
  ``wave.''
\begin{multicols}{4}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
\includegraphics{diagrams/thesis/product-one-wire.pdf}
}
\end{center}
\columnbreak
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
}
\end{center}
%% \end{multicols}
%% \begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
\includegraphics{diagrams/thesis/product-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
}
\end{center}
\end{multicols}

\item Sum types may similarly be represented by one wire or using
  parallel wires with a {{+}} operator between them. When tracing the
  execution of two additive wires, a value can reside on only one of the two
  wires.
\begin{multicols}{4}
%\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
\includegraphics{diagrams/thesis/sum-one-wire.pdf}
}

@@

\item Associativity is implicit in the graphical language. Three parallel
  wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
%% \begin{center}
%% \scalebox{0.95}{
%% %%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
%% \includegraphics{diagrams/thesis/assoc.pdf}
%% }
%% \end{center}

\item Commutativity is represented by crisscrossing wires.  By
  visually tracking the flow of particles on the wires, one can verify
  that the expected types for commutativity are satisfied.

@@

\item The morphisms that witness that {{0}} and {{1}} are the additive and
  multiplicative units are represented as shown below. Note that since there
  is no value of type 0, there can be no particle on a wire of type {{0}}.
  Also since the monoidal units can be freely introduced and eliminated, in
  many diagrams they are omitted and dealt with explicitly only when they are
  of special interest.
\begin{multicols}{4}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
\includegraphics{diagrams/thesis/uniti.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
\includegraphics{diagrams/thesis/unite.pdf}
}
\end{center}  
%% \end{multicols}
%% \begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
\includegraphics{diagrams/thesis/zeroi.pdf}
}
\end{center}
\columnbreak
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
\includegraphics{diagrams/thesis/zeroe.pdf}
}
\end{center}
\end{multicols}

\item Distributivity and factoring are interesting because they
  represent interactions between sum and pair types. Distributivity
  should essentially be thought of as a multiplexer that redirects the
  flow of {{v:b}} depending on what value inhabits the type {{b1+b2}},
  as shown below. Factoring is the corresponding adjoint operation.

@@

\item The {{trace+}} operation is drawn as a looped circuit, where the
  traced type {{b1}} is shown as flowing backward. 
%% The semantics and
%%  interpretation of {{trace}} are dealt with in detail in previous
%%  work (see Sec. 4 of \cite{James:2012:IE:2103656.2103667}).

@@

\paragraph*{Universality.} 
As an example, consider the implementation of the Toffoli gate below,
which establishes that {{langRevT}} is complete for combinational
circuits. This encoding uses the type {{1+1}} as {{bool}} and values
{{left ()}} and {{right()}} as {{true}} and {{false}}. Previous
work\cite{James:2012:IE:2103656.2103667} (see Sec 5) also establishes
that {{langRevT}} is Turing complete by showing the compilation of a
Turing complete language -- a first-order language with numbers and
iteration -- to {{langRevT}}.

\begin{center}
  \includegraphics{diagrams/toffoli.pdf}
\end{center}

@@

\noindent The machine is started with a number {{n}} in the first position
and {{0}} in the second. Each reduction step decrements the first number and
increments the second. The machine stops when the first number reaches 0,
thereby taking exactly {{n}} steps. For example, if the machine is started in
the configuration {{<3,0>}}, it would take exactly 3 steps to reach the final
configuration: {{ <3,0> |--> <2,1> |--> <1,2> |--> <0,3> }}. Dually, since
the machine transition is clearly reversible, the machine can execute
backwards from the final configuration to reach the initial configuration in
also 3 steps.

Our goal is to implement this abstract machine in {{langRevT}}: We start by
writing a combinator {{c : nat * nat <-> nat * nat}} that executes one step
of the machine transition. Here {{nat}} is an abbreviation for the recursive
type {{@@x.1+x}}. The combinator, when iterated, should map {{(n,0)}} to
{{(0,n)}} by precisely mimicking the steps of the abstract machine. Let us
analyze this desired combinator {{c}} in detail starting from its interface:

@@

The triangle on the left denotes the {{unfold}} isomorphism, which works as
follows: If the number {{n}} is zero, the output is the top branch which has
type {{1}}. If the number was non-zero, the output is on the bottom branch
and has value {{n-1}}. The triangle on the right is the dual {{fold}}
isomorphism.  We now use distribution to isolate each possible machine state:

@@

We are now close to the completion of the interpreter.  The branch labeled
{{((), m)}} corresponds to the machine state {{<0,m>}} which is the stop
state of the machine. Similarly the branch labeled {{((), n')}} corresponds
to the start state of the machine.  The last step in the construction
involves introducing a {{trace}} operation for iterating the small step
realized so far. We do this by first ``sliding'' the two sections of the
diagram as shown below:

@@

\noindent
we exchange sections A and B while maintaining the connectivity of the wires:

@@

This rearrangement exposes that the middle connections were really recursive
calls to the machine transition. The final figure is given below:

@@

\begin{center}
\scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
\end{center}

@@

\paragraph*{Input Machines.}

Our translation apply to machines presented as small-step operational
semantics, i.e. machines that are step-wise rewrites of tuples of
algebraic datatypes, wherein each rewrite step must be logically
reversible. The formal criterion logical reversibility in this setting
are captured in Abramsky's presentation of \emph{bi-orthogonal
  automata} \cite{abramsky2005structural} and may be summarized as:
\begin{enumerate}
\item
  Machines must have distinguishable start and stop states. There may
  be more than one valid start state and more than one valid stop
  state.
\item Each valid machine state must match a unique pattern on the
  left-hand-side and right-hand-side of the `{{|-->}}' rewrite.
\item Every reduction step must be computable (in {{langRevT}}) and
  must be logically reversible -- i.e. it must be possible to reduce
  from right to left.
\end{enumerate}

@@

\paragraph*{Steps of the Translation.}

Although trivial, the previous example captures the fundamental steps
of our construction. For reference, these steps are:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the left
  hand side of each machine state; 
\item Expand the output to expose enough structure to distinguish the right
  hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. This is always a
  mechanical process completely independent of the logic of the machine.
\end{enumerate}

\subsection{Adder} 
\label{sec:adder}

@@

The reduction rules on the right traverse a given tree and increment
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have a two types of machine
states -- {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored subtrees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

\paragraph*{New Notation.}
To make the diagrams easier to understand, we introduce a syntactic
convenience which combines the first steps of the construction that consist
of {{fold}} / {{unfold}} and {{distribute}} / {{factor}}. We collectively
represent these steps using thin vertical rectangle. Also we will introduce
the convention that the component that is being expanded (or constructed) will be
marked by using {{[^ ^]}} and the components that are being generated (or consumed) will be
marked by {{[v v]}}. For example, given a value of type {{tree*nat}}, the
diagram below shows how to first expand the {{tree}} component and then in
one of the generated branches, expand the {{nat}} component:

@@

%subcode{opsem} include main
%! columnStyle = rclr
% <true, L n, c> &|-->& <false, L (n+1), c> &~~~~~ rule 1
% <true, N t1 t2, c> &|-->& <true, t1, Lft c t2> & rule 2 
% <false, t1, Lft c t2> &|-->& <true, t2, Rgt t1 c> & rule 3
% <false, t2, Rgt t1 c> &|-->& <false, c, N t1 t2> & rule 4
\end{multicols}
\end{small}

And we start examining the machine components as before:

@@

On the input side, for {{true}} states, we have expanded the tree component and
for {{false}}~states we have expanded the tree contexts. Similarly we have done
the opposite on the output side exactly matching up what the abstract machine
does. One thing to note is that we dropped the {{1}} introduced by expanding
the {{bool}} and instead just labeled the {{true}} and {{false}} branches.

@@

\item The definition here shows only the handling of the composition
  combinators, which are the interesting cases. All the primitive
  isomorphisms are hidden in the {{iso(v)}} application in {{rule~1}}. This
  should be read as ``transform {{v}} according to the primitive isomorphism
  {{iso}}.'' Implementing this is straightforward and is hence skipped.

@@

>>>>

@@

\begin{document}
\title{Isomorphic Interpreters from Logically Reversible Abstract Machines} 
\titlerunning{On the Construction of Isomorphic Interpreters} 
\author{Roshan P. James \and Amr Sabry}
\institute{School of Informatics and Computing, Indiana University\\
\mails}
\maketitle

@@

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are given some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Axelsen:2011:RPC:1987171.1987176,Yokoyama:2008:PRP,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of transforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

\paragraph*{Graphical Language.}
Following the tradition for computations in monoidal
categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
graphical notation that conveys the semantics of {{langRevT}}.  The general
idea of the graphical notation is that combinators are modeled by ``wiring
diagrams'' or ``circuits'' and that values are modeled as ``particles'' or
``waves'' that may appear on the wires. Evaluation therefore is modeled by
the flow of waves and particles along the wires.

\begin{itemize}
\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
  is simply represented as a wire labeled by its type {{b}}, as shown below
  on the left. In more complex diagrams, if the type of a wire is obvious
  from the context, it may be omitted. When tracing a computation, one might
  imagine a value {{v}} of type {{b}} on the wire, as shown below on the
  right:

@@

\item The product type {{b1*b2}} may be represented using either one wire
  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
  case of products represented by a pair of wires, when tracing execution
  using particles, one should think of one particle on each wire or
  alternatively as in folklore in the literature on monoidal categories as a
  ``wave:''
\begin{multicols}{4}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
\includegraphics{diagrams/thesis/product-one-wire.pdf}
}
\end{center}
\columnbreak
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
}
\end{center}
%% \end{multicols}
%% \begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
\includegraphics{diagrams/thesis/product-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
}
\end{center}
\end{multicols}

\item Sum types may similarly be represented by one wire or using parallel
  wires with a {{+}} operator between them. When tracing the execution of two
  additive wires, a value can reside on only one of the two wires:
\begin{multicols}{4}
%\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
\includegraphics{diagrams/thesis/sum-one-wire.pdf}
}

@@

\item Associativity is implicit in the graphical language. Three parallel
  wires may represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}.
%% \begin{center}
%% \scalebox{0.95}{
%% %%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
%% \includegraphics{diagrams/thesis/assoc.pdf}
%% }
%% \end{center}

\item Commutativity is represented by crisscrossing wires:

@@

\item The morphisms that witness that {{0}} and {{1}} are the additive and
  multiplicative units are represented as shown below. Note that since there
  is no value of type 0, there can be no particle on a wire of type {{0}}.
  Also since the monoidal units can be freely introduced and eliminated, in
  many diagrams they are omitted and dealt with explicitly only when they are
  of special interest:
\begin{multicols}{4}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
\includegraphics{diagrams/thesis/uniti.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
\includegraphics{diagrams/thesis/unite.pdf}
}
\end{center}  
%% \end{multicols}
%% \begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
\includegraphics{diagrams/thesis/zeroi.pdf}
}
\end{center}
\columnbreak
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
\includegraphics{diagrams/thesis/zeroe.pdf}
}
\end{center}
\end{multicols}

\item Distributivity and factoring are interesting because they
  represent interactions between sum and pair types. Distributivity
  should essentially be thought of as a multiplexer that redirects the
  flow of {{v:b}} depending on what value inhabits the type {{b1+b2}},
  as shown below. Factoring is the corresponding adjoint operation:

@@

\item The {{trace+}} operation is drawn as a looped circuit, where the
  traced type {{b1}} is shown as flowing backwards:
%% The semantics and
%%  interpretation of {{trace}} are dealt with in detail in previous
%%  work (see Sec. 4 of \cite{James:2012:IE:2103656.2103667}).

@@

\paragraph*{Universality.} 
As an example, consider the implementation of the Toffoli gate below, which
establishes that {{langRevT}} is complete for combinational circuits. This
encoding uses the type {{1+1}} as {{bool}} and values {{left ()}} and
{{right()}} as {{true}} and {{false}}:
\begin{center}
  \includegraphics{diagrams/toffoli.pdf}
\end{center}
\noindent Previous work~\cite[Sec.~5]{James:2012:IE:2103656.2103667}
establishes that {{langRevT}} is Turing complete by showing the compilation
of a Turing complete language --- a first-order language with numbers and
iteration --- to {{langRevT}}.

@@

\noindent The machine is started with a number {{n}} in the first position
and {{0}} in the second. Each reduction step decrements the first number and
increments the second. The machine stops when the first number reaches 0,
thereby taking exactly {{n}} steps. For example, if the machine is started in
the configuration {{<3,0>}}, it would take exactly 3 steps to reach the final
configuration: {{ <3,0> |--> <2,1> |--> <1,2> |--> <0,3> }}. Dually, since
the machine transition is clearly reversible, the machine can execute
backwards from the final configuration to reach the initial configuration in
also~3 steps.

Our goal is to implement this abstract machine in {{langRevT}}. We start by
writing a combinator {{c : nat * nat <-> nat * nat}} that executes one step
of the machine transition. Here {{nat}} is an abbreviation for the recursive
type {{@@x.1+x}}. The combinator, when iterated, should map {{(n,0)}} to
{{(0,n)}} by precisely mimicking the steps of the abstract machine. Let us
analyze this desired combinator {{c}} in detail starting from its interface:

@@

The triangle on the left denotes the {{unfold}} isomorphism, which works as
follows: If the number {{n}} is zero, the output is the top branch which has
type~{{1}}. If the number is non-zero, the output is on the bottom branch
and has value {{n-1}}. The triangle on the right is the dual {{fold}}
isomorphism.  We now use distribution to isolate each possible machine state:

@@

We are now close to the completion of the interpreter.  The branch labeled
{{((), m)}} corresponds to the machine state {{<0,m>}} which is the stop
state of the machine. Similarly the branch labeled {{((), n')}} corresponds
to the start state of the machine.  The last step in the construction
involves introducing a {{trace}} operation for iterating the small step
realized so far:

@@

\noindent
Sliding sections A and B past each other while maintaining the connectivity
of the wires, exposes that the middle connections were really recursive calls
to the machine transition:

@@

%% The final figure is given below:

@@

%% \begin{center}
%% \scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
%% \end{center}

@@

@@

\paragraph*{Steps of the Construction.}

Although trivial, the previous example captures the fundamental aspects of
our construction. In general, our starting point is an abstract machine in
which every rewrite step is logically reversible. The formal criterion of
logical reversibility in this setting is captured by Abramsky's
\emph{bi-orthogonal automata}~\cite[Sec.~2]{abramsky2005structural} which is
summarized as:
\begin{enumerate}
\item
  Machines must have distinguishable start and stop states. There may
  be more than one valid start state and more than one valid stop
  state, however.
\item Each valid machine state must match a unique pattern on the
  left-hand side and right-hand side of the `{{|-->}}' relation.
\item Every reduction step must be computable (in {{langRevT}}) and
  must be logically reversible --- i.e. it must be possible to reduce
  from right to left.
\end{enumerate}
Given such a machine the process of encoding the machine in {{langRevT}}
consists of the following steps:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the
  left-hand side of each machine state;
\item Expand the output to expose enough structure to distinguish the
  right-hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. 
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Adder} 
\label{sec:adder}

@@

The reduction rules on the right traverse a given tree and increment
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have a two types of machine
states {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored subtrees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

\paragraph*{New Notation.}
To make the diagrams easier to understand, we introduce a syntactic
convenience which combines the first steps of the construction that consist
of {{fold}} / {{unfold}} and {{distribute}} / {{factor}}. We collectively
represent these steps using thin vertical rectangle. Also we will introduce
the convention that the component that is being expanded (or constructed) will be
marked by using {{[^ ^]}} and the components that are being generated (or consumed) will be
marked by~{{[v v]}}. For example, given a value of type {{tree*nat}}, the
diagram below shows how to first expand the {{tree}} component and then in
one of the generated branches, expand the {{nat}} component:

@@

%subcode{opsem} include main
%! columnStyle = rclr
% <true, L n, c> &|-->& <false, L (n+1), c> 
% <true, N t1 t2, c> &|-->& <true, t1, Lft c t2> 
% <false, t1, Lft c t2> &|-->& <true, t2, Rgt t1 c> 
% <false, t2, Rgt t1 c> &|-->& <false, c, N t1 t2> 
\end{multicols}
\end{small}

We start examining the machine components as before:

@@

On the input side, for {{true}} states, we have expanded the tree component
and for {{false}}~states we have expanded the tree contexts. We have done the
opposite on the output side exactly matching up what the abstract machine
does. One thing to note is that we dropped the {{1}} introduced by expanding
the {{bool}} and instead just labeled the {{true}} and {{false}} branches.

@@

\item The definition here shows only the handling of the composition
  combinators, which are the interesting cases. All the primitive
  isomorphisms are hidden in the {{iso(v)}} application in {{rule~1}}. This
  should be read as ``transform {{v}} according to the primitive isomorphism
  {{iso}}.''

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Acknowledgments.} 
This material is based upon work supported by the National Science Foundation
under Grant No. 1116725.
